Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-31 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: theory section and new refs

This theory belongs to the class of scalar-Gauss-Bonnet theories, described by
the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g = \textrm{det} \, g$ is the metric determinant, $R$ is the Ricci
scalar and $\varphi$ is a dynamical, massless scalar field that couples to the
Gauss-Bonnet invariant
%
% \begin{equation} \label{eq:def_gb}
$ \mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
% \end{equation}
%
where $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are respectively the Riemann and Ricci tensors.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
It can however contribute to the field equation when coupled to $\varphi$, as
done through function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi) = \dd f  / \dd \varphi$
is always nonzero and BHs are known to always support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which $f = \exp(\varphi)$.
%
% Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
% $f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the quadratic $f \propto \varphi^2$~\cite{Silva:2017uqg}
and Gaussian $f \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} coupling functions.

BHs in both classes of theories have hair, to which we can associate a (monopole) scalar charge,
related to the asymptotic $1/r$ decay of the scalar field.
%
This charge is not an independent parameter, and depends on the BH's mass and spin,
thus being a ``secondary charge''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is coupled to a curvature invariant, the scalar charge is larger (smaller)
the smaller (larger) the BH's mass is.

These properties are not mere theoretical curiosities, but have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar dipole radiation~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo} making
the theory prone to be constraint with present
GWs observations of BH binaries (BBH)~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
Second, the scalar field does influence the response of BHs to linearized perturbations in this theory,
affecting their QNM spectra.

The calculation of the QNMs in this theory is more complicated than in GR already for
nonrotating BHs. The reason is due to the coupling between scalar field and
the Gauss-Bonnet invariant, which manifests into a coupling between scalar perturbation
and gravitational perturbations of polar parity.
%
BH QNMs in EdGB gravity were first computed in Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn} and these calculations
were extended to leading-order in BH spin effects in Ref.~\cite{Pierini:2021jxd}.

@@

This theory is given by the following action,
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudo-scalar field, which couples
to the Pontryagin density,
%
\begin{equation}
    {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
\end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\tfrac{1}{2} \epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma}$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The coupling between scalar field and the Pontryagin density is set by the
$\ell_{\rm CS}$ with dimensions of length.

>>>>

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g = \textrm{det} \, g$ is the metric determinant, $R$ is the Ricci
scalar and $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
% \begin{equation} \label{eq:def_gb}
$ \mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
% \end{equation}
%
where $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are respectively the Riemann and Ricci tensors.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
It can however contribute to the field equation when coupled to $\varphi$, as
done through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi) = \dd f  / \dd \varphi$
is always nonzero and BHs are known to always support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which $f(\varphi) = \exp(\varphi)$~\cite{Kanti:1995vq}.
%
% \begin{equation}
% f = \exp(\varphi)\,,
% \end{equation}
%
%
% Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
% $f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Macedo:2019sem,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the Gaussian $f(\varphi) \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} and
the quadratic $f(\varphi) \propto \varphi^2$~\cite{Silva:2017uqg} coupling functions.

BHs in EdGB gravity have scalar hair, to which we can associate a (monopole) scalar charge $Q$,
related to the asymptotic $1/r$ fall-off of the scalar field.
%
This charge is not an independent parameter, and depends on the BH's mass and spin, %~\cite{Berti:2018cxi},
thus being a ``secondary hair''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is coupled to a curvature scalar, the $Q$ is larger (smaller)
the smaller (larger) the BH's mass is.

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar dipole radiation (see e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GWs
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the neutron star-black hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm GB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm GB}$ from a selection of black hole binaries (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.
%
% These works use the \emph{inspiral} portion of GW event only and a constraint
% on EdGB gravity can be obtained because black holes in this theory carry a
% monopole scalar charge which in turn can source dipolar scalar radiation,
% that affects the GW phase at $-1$PN order, relative to the dominant GR contribution, and
% is proportional to the difference between the charges of binary components.

Second, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH's QNM spectra.
%
% The calculation of the QNMs in this theory is more complicated than in GR already for
% nonrotating BHs. The reason is due to
The coupling between scalar field and the Gauss-Bonnet invariant, which
manifests into a coupling between scalar perturbation and gravitational
perturbations of polar parity, which, e.g., breaks the familiar equivalence
between the QNM spectra of polar and axial gravitational perturbations of
Schwarzschild BHs in GR.
%
The QNMs on nonrotating BHs in EdGB gravity were first computed in
Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn} and these calculations were
extended to leading-order in BH spin in Ref.~\cite{Pierini:2021jxd}.

@@

This theory is described by the action~\cite{Jackiw:2003pm},
%
\begin{equation} \label{eq:action_dcs}
    S_{\rm CS} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \vartheta)^2
    + \tfrac{1}{4} \ell^{2}_{\rm CS} \, \vartheta \, {}^{*}RR
    \right],
\end{equation}
%
where $\vartheta$ is a pseudo-scalar field, which couples
to the Pontryagin density
${}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma}$,
%
% \begin{equation}
%     {}^{*}RR = {}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} R^{\nu}_{\mu\rho\sigma},
% \end{equation}
%
where ${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma}$ is the dual of the Riemann tensor,
defined as
%
${}^{*}R^{\mu}{}_{\nu}{}^{\rho\sigma} =
\epsilon^{\mu}{}_{\nu\gamma\delta}
R^{\gamma\delta\rho\sigma} / 2$,
%
and where $\epsilon^{\mu}{}_{\nu\gamma\delta}$ is the Levi-Civita tensor.
%
The coupling strength between scalar field and the Pontryagin density is set by the
coupling constant $\ell_{\rm CS}$, with dimensions of length.
